---
title: "SSRP 2015"
author: "Lennon Lab"
date: "Monday, July 13, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---


## OVERVIEW
Here, we will use R and RStudio to examine patterns of change in resource use between an ancestral strain a derived strain. We will begin by focusing on three aspects:

### 1.) Average resource use 
One of the simplest, and hence, first questions to ask is whether there are directional changes in average resource use and the average number of resources used. Does the derived use more resources or less resources? Is the derived able to use each resource to a greater or lesser degree?

### 2.) Resource breadth vs. capacity
A second and more in-depth question is how the number of resources used relates to the degree to which an organisms typically uses resources. This may relate to the potential for the organism to use generalist (i.e., okay at using many resources, "jack of all trades") and specialist strategies (i.e., really good at using just a few resources).

### 3.) Changes in the metabolic "fingerprint"
Note how the first two questions do not actually account for *which* resources get used or how the various resources are related by a common element or nutrient. However, by combining information on which resources are used, how greatly they are used, and how the resources are related, we can obtain a metabolic "fingerprint" for the organism. In this way, we can ask whether the fingerprint has changed between the ancestral strain and the derived strain.


## 1.) SETUP
### A. Clear the Workspace and Retrieve and Set the Working Directory
We will begin our analysis by clearing our workspace of any variables, that is, leftover information from any previous work. We then set our 'working directory', which points R to the folder that contains our project's files (i.e., SSRP_2015).

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
rm(list = ls()) 
getwd() 
setwd("~/GitHub/SSRP_2015") 
```

### Install Packages
Next, we will import statistical packages that will allow us to analyze our data with specific tools.
We will begin by importing a package called `vegan`, which is often used by biologists and ecologists.
Let's install the `vegan` package and its dependencies. 
If you are running `vegan` for the first time, you will need to install it with the `install.packages` function. 
Otherwise, we recommend you just use the `require` function to load `vegan` and its dependencies.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
#install.packages("vegan") 
#install.packages("reshape") 
require("vegan")
require("reshape")
```

## 2) LOADING DATA
Next, let's load the data generated from our plate reader and the Ecolog software as well as a table of names corresponding to each resource and location on the plates.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
file.path <- "~/GitHub/SSRP_2015/data/cleaned/"
dat.frame <- data.frame(date = integer(), strain = character(),
                 AorD = character(), plate = character(),
                 plate.replicate = character(), duration = character())
  
  
file.names <- list.files(path = file.path, all.files=FALSE, 
                         full.names=FALSE, recursive=FALSE,
                         ignore.case=FALSE, include.dirs=FALSE)

for (name in file.names){
  print(name)
  file.name.info <- strsplit(paste(file.path, name, sep=''), "\\.") # split file name  
  print(file.name.info)
  date <- file.name.info[[1]]
  strain.id <- file.name.info[[2]]
  AorD <- file.name.info[[3]] # ancestral or derived
  plate <- file.name.info[[4]]
  plate.replicate <- file.name.info[[5]]
  duration <- file.name.info[[6]]
  #dat <- as.matrix(read.table(paste('~/GitHub/SSRP_2015/data/cleaned/', name))
  }



derived <- as.matrix(read.table("~/GitHub/SSRP_2015/data/cleaned/0723_3C_07_12_2015_RNF_parsed.txt"))

resource.names <- as.matrix(read.table("~/GitHub/SSRP_2015/ecoplate.info/resource_matrix.txt"))


file.names
mol.groups <- matrix("NA", nrow = 8, ncol = 12)
mol.groups <- as.matrix(mol.groups)
```

### 3) NORMALIZING DATA
Because the number of cells can vary across plates, it's important to normalize our data, i.e., make our resulting values directly comparable. We will do this using the value recorded for the well where there was only **water** from the ancestral plate.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
# We reshape our data from a matrix to long format
m1.a <- melt(ancestral); m1.d <- melt(derived)
m1.names <- melt(resource.names); m1.groups <- melt(mol.groups)

# Next, merge the columns of interest together
dat <- data.frame(mol.groups[,3], as.character(m1.names[,3]), m1.a[,3], m1.d[,3])

# Remove the background
background <- 0.042

dat[,3:4] <- dat[,3:4] - background
colnames(dat) <- c("sub.group","substrate","anc","der")

# Calculate the mean reading for water
anc.water <- mean(subset(dat, dat[,2]=="Water")[1:3,3])
der.water <- mean(subset(dat, dat[,2]=="Water")[1:3,4])

#Remove background and Normalize the data
dat$anc.norm <- dat$anc * (anc.water/anc.water) 
dat$der.norm <- round(dat$der * (anc.water/der.water), digits = 3) 
```


### 1.) Average resource use
Let's examine the following questions:
1) How many resources were used by each strain?
2) What resources are used by each strain?

Because water does not contain a carbon substrate, it will act as our control data. If the adsorbance value is greater than the adsorbance of water, we will include it in the number of used resources.

```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
m2 <- melt(dat, ids = c("sub.group","substrate"))
c1 <- cast(m2, substrate ~ variable , mean)

threshold <- 0.037

length(c1[,3][c1[,3] > threshold])
length(c1[,4][c1[,4] > threshold])
```

Let's visualize the resources used per isolate
```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
c1 <- as.matrix(c1)

barplot(c1[,3:4],
        width = 1, beside = T,
        ylim = c(0,1), las = 1, cex.axis = 1.15,
        main = "Resource Use",
        xlab = list("Substrate", cex = 1.5, font = 2),
        ylab = list("OD", cex = 1.5, font = 2))

```


```{r, results = 'hide', echo=TRUE, message = FALSE, warning = FALSE}
# Demonstration of a scatter plots 
x <- c(1,2,3,4,5,6,7,8,9)
y <- c(1,2,3,4,5,6,7,8,9)

plot(x, y, 
     xlim = c(min(x),max(x)), ylim = c(min(y),max(y)), # specifiy your plot range
     xlab = list("x", cex = 1.5, font = 2),  # label the x-axis
     ylab = list("y", cex = 1.5, font = 2),  # label the y-axis            
     cex = 2, cex.lab = 1.5, cex.axis = 1.15 # change the sizes
     )
```